Climate indicators

### 2023-02-24 DRAFT_climate_indicators.Rmd
switch to st_apply

```{r}
microOut <- microbenchmark(
st_apply =  x_aug %>%
  filter(band ==1) %>% # for our real data we don't need this filter
  st_apply(1:2, mean, rename=F),
aggregate = x_aug %>%
  aggregate(by = "year", mean),
times=30
)
```

@@

`st_apply` is considerably faster, even wit the filtering step, but `aggregate` has the advantage of that it retains the time dimension. 
I think I will try to use `aggregate`, and it it takes too long on the real data, then I can come back to this point, and perhaps switch to st_apply.

```{r}
x_summerMean <-   x_aug %>%
  filter(band ==1) %>% 
  st_apply(1:2, mean, rename=F)
```

@@

ggplot() + 
  geom_stars(data = x_summerMean_v2) +
  coord_equal() +
  facet_wrap(~time) +
  theme_void() +
  scale_fill_viridis_c(option = "D") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))
)
```

@@

First, let me cut the _band_ dimension.
```{r}
temp <- x_summerMean_v2 %>%
  filter(band==1) %>%
  adrop(drop=4) %>%
  aperm(c(2,3,1))
dim(temp)
```

```{r}
oldTime <- st_get_dimension_values(x_summerMean_v2, "time")

# Function to add random noise
myRandom <- function(x) x*rnorm(1,1,.05)

y1 <- temp
y2 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
y3 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
y4 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
y5 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)

# We combine the data into one cube:
y <- c(y1, y2, y3, y4, y5)
```

```{r}
ggplot() + 
  geom_stars(data = y) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~time)
```

Now we aggregate across years. For the reference period is is probably best to use *Method 1*, and for he indicator variable (the moving window estimation) probably *Method 2* is best. So Method 1 now.

@@

```{r}
y_ref <- setNames(y_ref, c("reference_value", "sd"))
```

```{r}
tmap_arrange(
tm_shape(y_ref)+
  tm_raster("reference_value")
,
tm_shape(y_ref)+
  tm_raster("sd",
            palette = "-viridis")
)
```

@@

```{r}
y_var <- y %>% mutate(Attribute_A = Attribute_A * 0.98) %>% 
  split(3) %>% c(y_ref)
y_var <- y_var %>%
  setNames(paste0("v_",substr(names(y_var), 1,4)))%>%
  rename("reference" = v_refe,
         "sd" = v_sd)
```

```{r}
# select the columns to normalise
cols <- names(y_var)[!names(y_var) %in% c("reference", "sd") ]
cols_new <- cols
names(cols_new) <- gsub("v_", "i_", cols)

# Mutate
# can I find a way to keep the v_ columns?
y_var_norm <- y_var %>%
  mutate(lower = reference - 5*sd ) %>%
  mutate(threshold = reference -2*sd ) %>%
  mutate(across(all_of(cols), ~ if_else(.x < threshold, 
                                        (.x - lower) / (threshold - lower),
                                        (.x - threshold) / (reference - threshold),
                                        ))) %>%
  mutate(across(all_of(cols), ~ if_else(.x > 1, 1, .x))) %>%
  mutate(across(all_of(cols), ~ if_else(.x < 0, 0, .x))) %>%
  rename(all_of(cols_new))
```

```{r}
ggarrange(
ggplot() + 
  geom_stars(data = y_var["v_1970"]) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))
,
ggplot() + 
  geom_stars(data = y_var_norm["reference"]) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))
,
ggplot() + 
  geom_stars(data = y_var_norm["i_1970"]) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))
, ncol=3
)
```

#### Normalise variable 
I now want to do cell-by-cell computations across the _attributes_ dimension, and divide _Attribute_A_ by _reference_value_.

```{r}
# A function for normalization
# I need to add a break point and lower reference value here as well.
# And I need to add truncation.
myFUN <- function(a,b) (a/b)
```

```{r}
system.time(
  temp <- st_apply(y_v_merged, 
                   c("x", "y"),
                   FUN = myFUN
  )
)

```
This operation is quite fast!

```{r, fig.cap = "The normalised indicator for year 1970."}
tm_shape(temp)+
  tm_raster()+
  tm_layout(legend.outside = T)
```
This looks like perfect noise, and it is, because the variable and reference value are so well correlated.
If the reference value was static, it would look different.

```{r}
y_v_merged$value[,,2] <- 200
```

```{r}
ggplot() + 
  geom_stars(data = y_v_merged) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~attributes)
```

```{r}
temp <- st_apply(y_v_merged, 
                   c("x", "y"), 
                   #1:2,
                   FUN = myFUN
)
```

```{r, fig.cap = "The normalised indicator for year 1970 when the reference value is the same all over."}
tm_shape(temp)+
  tm_raster()+
  tm_layout(legend.outside = T)
```

>>>>

```{r}
temp_names <- year(st_get_dimension_values(x_aug, "time")[1])

microOut <- microbenchmark(
st_apply =  x_aug %>%
  filter(band ==1) %>% # for our real data we don't need this filter
  st_apply(1:2, mean) %>%
  setNames(paste0("v_",temp_names)),
aggregate = x_aug %>%
  aggregate(by = "year", mean),
times=30
)
```

@@

`st_apply` is considerably faster, even wit the filtering step, but `aggregate` has the advantage of that it retains the time dimension, whereas for `st_apply` I need to set the attribute name to be the year. I will try to use `st_apply`.

```{r}
temp_names <- year(st_get_dimension_values(x_aug, "time")[1])
x_summerMean <-   x_aug %>%
  filter(band ==1) %>% 
  st_apply(1:2, mean) %>%
  setNames(paste0("v_",temp_names))
```

@@

ggplot() + 
  geom_stars(data = x_summerMean) +
  coord_equal() +
  #facet_wrap(~time) +
  theme_void() +
  scale_fill_viridis_c(option = "D") +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))
)
```

@@

First, let me cut the _band_ dimension.
```{r}
#temp <- x_summerMean %>%
#  filter(band==1) %>%
#  adrop(drop=4) %>%
#  aperm(c(2,3,1))
#dim(temp)
```

```{r}
#oldTime <- st_get_dimension_values(x_summerMean_v2, "time")
#
## Function to add random noise
#myRandom <- function(x) x*rnorm(1,1,.05)
#
#y1 <- temp
#y2 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-1)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
#y3 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-2)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
#y4 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-3)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
#y5 <- temp %>% st_set_dimensions("time", values = oldTime %m+% years(-4)) %>% st_apply(c("x", "y", "time"), FUN = myRandom)
#
#
## We combine the data into one cube:
#y <- c(y1, y2, y3, y4, y5)
```

```{r}
# Function to add random noise
myRandom <- function(x) x*rnorm(1,1,.05)
```

```{r}
y1 <- x_summerMean
y2 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-1)) 
y3 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-2)) 
y4 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-3)) 
y5 <- x_summerMean %>% st_apply(1:2, myRandom) %>% setNames(paste0("v_",temp_names-4)) 
# We combine the data into one cube:
y <- c(y1, y2, y3, y4, y5) %>% 
  merge(name = "Year") %>%
  setNames("Attribute_A")
```

```{r}
ggplot() + 
  geom_stars(data = y) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))+
  facet_wrap(~Year)
```

Now we aggregate across years, again using st_apply.

@@

```{r}
y_ref <- setNames(y_ref, c("reference_upper", "sd"))
```

```{r}
tmap_arrange(
tm_shape(y_ref)+
  tm_raster("reference_upper")
,
tm_shape(y_ref)+
  tm_raster("sd",
            palette = "-viridis")
)
```

@@

```{r}
y_var <- y %>% mutate(Attribute_A = Attribute_A * 0.98) %>% 
  split(3) %>% c(y_ref)
```

#### Normalise variable

```{r}
# select the columns to normalise
cols <- names(y_var)[!names(y_var) %in% c("reference_upper", "sd") ]
cols_new <- cols
names(cols_new) <- gsub("v_", "i_", cols)
```

```{r}
# Mutate
# can I find a way to keep the v_ columns?
system.time(
y_var_norm <- y_var %>%
  mutate(lower = reference_upper - 5*sd ) %>%
  mutate(threshold = reference_upper -2*sd ) %>%
  mutate(across(all_of(cols), ~ if_else(.x < threshold, 
                                        (.x - lower) / (threshold - lower),
                                        (.x - threshold) / (reference_upper - threshold),
                                        ))) %>%
  mutate(across(all_of(cols), ~ if_else(.x > 1, 1, .x))) %>%
  mutate(across(all_of(cols), ~ if_else(.x < 0, 0, .x))) %>%
  rename(all_of(cols_new)) %>%
  c(select(y_var, all_of(cols)))
)
```

I'm not sure mutate-across is the fastest way to do this. Perhaps st_apply would be better.

```{r}
ggarrange(
ggplot() + 
  geom_stars(data = y_var_norm["v_1970"]) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))
,
ggplot() + 
  geom_stars(data = y_var_norm["reference_upper"]) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))
,
ggplot() + 
  geom_stars(data = y_var_norm["i_1970"]) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c(option = "D") +  
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_discrete(expand = c(0, 0))
, ncol=3, align = "hv"
)
```

This looks like perfect noise, and it is, because the variable and reference value are so well correlated.
If the reference value was static, or if there was a pattern in v_1970, the scaled indicator map would look different.

#### Plot time series

```{r}
ts <- y_var_norm %>%
  select(all_of(starts_with("i_"))) %>%
  st_apply(MARGIN = 3, mean)
```
